Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2020-05-20 doc/response.tex

\revtext{
Lines 59, 114-115, 129, etc.: There is a distinction between "truncation"
and "censoring" in survival analysis. In the current context (e.g. contact tracing), when the inclusion criteria is conditioning on the infection
event of the infectee and looking backward to ascertain the infection time
of its infector, it is right-truncation. It leads to over-sampling short durations (as stated in Line 129). I suggest to use "right-truncation" instead
of "censoring" throughout the paper, wherever it appears.
}

We now use truncation instead of censoring.

\revtext{
Lines 121 and 123: The backward generation-interval distribution $b_t(\tau)$ and
the "censored" (right-truncated?) generation-interval distribution $c_t(\tau)$
should be clearly defined and expressed (e.g. in terms of conditional distribution) before expressing the relationship (6).
}

We now define both backward and forward generation-interval distributions more appropriately in the text.
We no longer use the term ``censored'' generation-interval distribution. 
We refer to it as the aggregated generation-interval distribution, which is a distribution of all realized generation intervals until certain time point.
We think thie definition provides a clearer definition.

@@

\revtext{
Lines 233-241: Once more, the sampling bias corresponds to right-truncation,
not censoring. A great deal of words is involved leading to gef $g_{eff}(\tau)$. The
mathematical language is more powerful and perhaps gef $g_{eff}(\tau)$ could be
explained more concisely (and take less space?)
}

We have removed this paragraph and tried to make the explanation more concise.
We no longer use the term effective generation-interval distribution but rather refer to it as the initial forward generation-interval distribution.

\revtext{Lines 241-243: "In a large homogeneously mixing population, the effective
distribution is expected to be equivalent to the intrinsic distribution..."
Define both $g(\tau)$ and gef $g_{eff}(\tau)$ mathematically at the first place and prove
their equivalence in a large homogeneously mixing population.}

We now show that the initial forward generation-interval distribution is equivalent to the intrinsic distribution.
We also point to the analytical intractability of the spatial effects on the realized generation-interval distributions (and therefore the initial forward distribution) and therefore cannot be expressed mathematically (for non-homogeneous cases):

``Since spatial effects on realized generation intervals are generally expected to be analytically intractable, even in simple networks (e.g., see \cite{tomba2010some} for discussion regarding the realized generation intervals in a household with one infector and two susceptibles), we rely on simulations hereafter to vadliate our prediction.''

\revtext{However, I suggest the authors to compare the "individual-based
method" in the paragraph starting from Line 252 to Line 260 with the statistical methods for analyzing right-truncated survival data as mentioned
above, and explicitly write out the likelihood function (Line 258) to see if
they are the same or different.}

The likelihood function is provided in the Methods section.
\swp{TODO. It seems like they're different... Ugh. I'll get back to this.}

\revtext{The authors has used the notations $\RR$ (Line 35) and $\Ro$ (Lines 98-99) for reproduction numbers and
linked them via $\RR = \Ro S(t)$:}

\revtext{Shouldn't $\RR$ be a function of time $t$ via $S(t)$?}

We no longer use $\RR$ to avoid confusion. We now use $\Rini = \Ro S(0)$ to reflect the possibility that $S(0)$ may not be 1 but clearly convey the idea that we're not estimating time dependent reproduction number.

\revtext{Is $S(t)$ a number or a proportion? In deterministic models, this matters
less as one can argue that all deterministic models are describing the
change of proportions (by re-scaling on population size, assuming it is
constant); but it matters a lot in stochastic models because the population
size is an essential parameter.}

We make these distinctions clear in the main text.

\revtext{
When stating that ``$\RR$'' transcends to the final size of the epidemic (Line
37), do the authors actually mean $\Ro$; which does not depend on time?
}

Yes. We now make a clear distinction between $\RR$ and $\Ro$.

\revtext{
When talking about the final size, one considers that the system is at a
disease-free equilibrium. When using the final size equation to calculate a
value for the reproduction number, the calculated value does not depend
on time. If there are behavior changes and public health interventions
throughout the outbreak, the calculated value $\RR_{final}$ is usually smaller
than $\Ro$ (defined at another disease-free equilibrium at the very moment
when an infector was seeded). So now we have three reproduction numbers, $\Ro$, a time-dependent $\RR = \Ro S(t)$ and $\RR_{final}$: Need to clarify the differences and reveal the hidden assumptions when discussing (and using)
each of them.
}

\swp{TODO? Just say we are interesting in the initial phase? maybe write something in the discussion? We're only using $\Rini$}

\revtext{
The authors stated that $\RR$ transcends to the final size of the epidemic
(with cited references). I boldly assume that the authors actually meant
by $\Ro$: This statement is in the context of a large outbreak that ends
when either there is no infectious individuals or there is no susceptible
individuals left in the population. Under a different context, the outbreak
may reach an endemic steady state with sustained on-going transmission
persisting for a very long period of time due to re-infection or demography.
It could be further said that $\Ro$ can transcend to the long term endemicity
level, such as $S(\infty) = \RR$: I leave to the authors to search the literature
for the latter
}

Yes, we meant $\Ro$. We have now clarified this in the text.
We now mention in Introduction that $\Ro$ can also affect the endemicity level and added a few citations.

\revtext{$g(\tau)$ as a p.d.f. given by (2) and Line 101 is also an equilibrium
distribution.}

While we agree with the reviewer that the intrinsic generation-interval can be interpreted an equilibrium distribution, this only holds under homogeneous mixing assumption. 
As we show in this paper, when there is spatial structure, the realized generation-interval distributions differ from the intrinsic one. 
In particular, even when the entire population is susceptible, the forward generation-interval distribution of the single primary infector will still be subject to the egeocentric effect.
Therefore, the realized generation intervals of a single, primary infector in an otherwise susceptible population will not follow the intrinsic generation-interval distribution. We mention in the manuscript that the intrinsic distribution can be interpreted as an equilibrium distribution but do not want to emphasize on using the term ``equilibrium'' particularly given that we're interested in realized distributions on a spatially heterogeneous network.
The reviewer also presented a series of formulas on the mean generation interval for special cases of the SEIR model without any suggestions;
given that our goal is not to explore all special cases of the SEIR model, we do not think presenting all those formulas is relevant to the discussion of the paper.

\revtext{
Expression (13) is the convolution (the sum of) the exponentially
distributed latent period with mean $1/\sigma$ and another exponential distribution, with mean $1/(\gamma+\lambda)$; as written in Lines 188-189. It is also equivalent
to the homogeneous mixing situation, in which, individuals during their
infectious periods are isolated (with rate $\lambda$) so that the "effective" duration
of the infectious period is exponentially distributed with mean $1/(\gamma + \lambda)$:
Any comments from the authors on this coincidence?
}

We explain this in the main text:

``In this case, with fixed infectiousness during the infection period, the effect of accounting for pairwise contacts is the same as an increase in the recovery rate (by the amount of the per-pair contact rate $\lambda$). 
Infecting a susceptible contact is analogous to recovery because the contactee cannot be infected again --- the infector cannot no longer transmit infection even if they are infectious (effectively losing infectiousness).
Therefore, the resulting egocentric generation-interval distribution is equivalent to the intrinsic generation-interval distribution with mean latent period of $1/\sigma$ and mean infectious period of $1/(\gamma + \lambda)$.''

\revtext{Under suitable assumptions on homogeneous mixing, the intrinsic
generation interval distribution $g(\tau)$ measures the infectiousness within
an infected individual without involving another individual. (In SEIS,
SEIR, SEIRS settings, it is a function of the latent period distribution
and the infectious period distribution.)}

We now define the intrinsic generation-interval distribution as follows as we think it is more concrete; it is also easier to distinguish intrinsic vs. realized distributiosn under this definition:

@@

``The \emph{intrinsic} generation-interval distributions describe the expected time distribution of infectious contacts made by a primary case.''

\revtext{The distribution $g(\tau)$ links the two intrinsic parameters R0 and r
(homogeneous mixing). Both $\Ro$ and $g(\tau)$ are intrinsic in the context of the equilibrium conditions. Therefore the parameter $r$ is a theoretical value and is very different from any observed exponential
or sub-exponential growth rate as the system is not at equilibrium.}

\swp{???}

\revtext{Observed exponential or sub-exponential growth events are different from the infection events. The current paper only addresses the infection events (mostly unobservable). This was
already addressed in the comments with respect to Line 47-48.
Any attempt to link the estimated $r$ based on data collected
on observed events using the Euler-Lotka equation could be extremely misleading}

It is not \emph{extremely} misleading to use the estimated $r$ from the observed data. 
See \swp{How do I respond to this, actually?}:

@@

``Ma, Junling, Jonathan Dushoff, Benjamin M. Bolker, and David JD Earn. "Estimating initial epidemic growth rates." Bulletin of mathematical biology 76, no. 1 (2014): 245-260.''

\revtext{
The intrinsic parameters R0 and r are parameters in branching
process, which is used to approximate the "initial phase" of the
outbreak and that approximation is poor. By the way, the condition in Lines 107-108 $S(t) = S$ is also inaccurate. The expected exponential growth is under the approximation $S(t)/n \approx 1$ and
$n \to \infty$; where $n$ is the population size. The so-called initial
phase (when the depletion of the susceptible population is negligible) totally ignores the difference between the initial disease free equilibrium and the highly explosive dynamic growth
}

``The expected exponential growth is under the approximation $S(t)/n \approx 1$ and
$n \to \infty$'' 
This is inaccurate.
It is still possible for exponential growth to occur when the population is not fully susceptible ($S(t) \approx S(0) \neq 1$) as long as the depletion of the susceptible population is negligible.
To avoid confusion, we now use the term ``initial reproduction number'': $\Rini = \Ro S(0)$.

\revtext{
 It is too strong a statement that "the Erlang distributions are
more realistic than the exponentially distributed periods in the standard
SEIR model." The advantage of using the Erlang distributions is that the
system can be analyzed using ordinary differential equations. They are
only suitable for latent and infectious periods with very small variances
}

The exponential distribution is a special case of the Erlang distribution (when the shape parameter is equal to 1).
We disagree the statement that it is ``too strong'' to say that a more general distribution is more realistic than its special case.
To avoid confusion, we now say refer to Erlang distributions as ``more flexible''.

\revtext{
The Erlang distribution is not only a subset of the gamma distribution with integer valued shape parameters, but also it is limited to distributions with variances smaller than the corresponding exponential distribution with the same mean value.
}

@@

The Erlang distribution includes the exponential distribution.

\revtext{
One can not even claim the gamma distributions are more realistic
without supported by data. For one thing: all gamma distributions
have exponential tails which may be contested in real data.
}

@@

One can at least claim that the gamma distribution is more realistic than its subset, the exponential distribution.

\revtext{
"... based on more realistic distributions in Appendix." I do
not see "more realistic distributions" more than the Erlang distributions.
How about heavy tailed distributions and distributions with large variances? In fact, many of the discussions in this paper can be generalized
to arbitrarily distributed latent and infectious periods.
}

We had omitted a sentence previous --- we have now clarified this sentence as follows:

``In the main text, we present results based on exponentially distributed latent and infectious periods;
we show results based on more Erlang distributed latent periods ($n_E=2$), which better matches the incubation period distribution of Ebola virus disease, in Appendix.''

\revtext{
It is more convenient to use the Laplace transforms as a tool.
}

The SEIR model is just an example --- the purpose of this paper is not to go through all possible generalizations of the SEIR model, but rather to explore spatial effects.
The Laplace tranforms are only more useful for the intrinsic generation-interval distributions under the homogeneous assumption. 
The forward generation-interval distribution in a spatially hetergeneous environment is analytically intractable and therefore, the Laplace transforms no longer hold.
Given that our goal is to infer the initial forward generation-interval distribution, rather than the intrinsic distribution, we feel that the Laplace transforms are not necessarily helpful for the discussion.
In particular, the reviewer asks us to prove special cases of the Euler-Lotka equation, which has been proven previosuly as suggested by the references, using Laplace transforms; 
we do not think this is relevant to the discussion of our paper.

\revtext{How to assess if the gamma distribution is a good choice as a
model? Run some kind of goodness-of-fit tests?}

The gamma distribution has been previously tested for many diseases. 
In our case, Figure 4 demonstrates that the gamma approximation of the generation-interval distribution provides a reasonable estimate for the initial reproduction number from the exponential growth rate.
We discuss potential caveats in assuming a shape of the generation-interval distribution in Discussion.

``We compare two methods for estimating the initial forward generation-interval distribution and assume that the initial forward generation-interval distribution follows a gamma distribution.
The gamma approximation of the generation interval distribution has been widely used due to its simplicity;
we previously thought that summarizing the entire distribution with two moments (mean and variance) is sufficient to understand the role of generation-interval distributions in linking $r$ and $\Rini$.
However, further investigation of our methods suggests that making a wrong distributional assumption can lead to biased estimates of the mean and CV of a generation-interval distribution (Appendix A.3), even though the gamma distribution ``looks'' indistinguishable from the shape of the intrinsic generation-interval distribution (derived from the SEIR model).
These results are particularly alarming because it is impossible to know the true shape of the generation-interval distribution for real diseases.
Nonetheless, biases in the parameter estimates of a generation-interval distribution may cancel out and give unbiased estimate of the reproductive number (Appendix A.3).''

\revtext{How does the likelihood
function look like? Since the "observed" generation-intervals (assuming
observable) do not arise from a designed random experiment, but observational data, I expect the likelihood function should reflect the observation
scheme (e.g. retrospective)? Please be explicit.
}

The population-level method does not take into account the observation scheme or the infection process. 
We simply fit the gamma distribution to all realized generation intervals (i.e., the aggregated distribution);
we feel that writing down the likelihood function is an overkill for a basic gamma fit.
We have clarified the text:

``The \textbf{population-level method} estimates the initial forward generation-interval distribution by reversing the inverse exponential weighting in the aggregated generation-interval distribution without explicitly accounting for the infection process (i.e., who infected whom):
\begin{equation}
f_0(\tau) \propto a_0(\tau) \exp(r\tau).
\end{equation}
In order to do so, we first approximate the aggregated distribution $a_0$ with a gamma distribution by assuming that aggregated realized generation intervals (subject to right-truncation) during the exponential growth phase come from the same gamma distribution;
specifically, we estimate the mean $\bar G$ and shape $\alpha$ of a gamma distribution by maximum likelihood.
Then, the initial forward generation-interval distribution also follows a gamma distribution with mean $\alpha/(\alpha/\bar G - r)$ and shape $\alpha$.
We then use the estimated initial forward generation-interval distribution to infer the initial reproductive number $\Rini$ from the estimated growth rate (using the Euler-Lotka equation).''

@@

\rev{
Elaborate how the estimation of $\RR$ is expected to be sensitive to under-reporting (of which data)
}

That was a bad notation on our side --- we should not have used $\RR$ in the likelihood. 
We have changed the notation for the Poisson process and provide a better explanation.

@@

\rev{
How exactly is $\RR$ (and, is it \Ro or some kind of time-dependent \RR) estimated in this paper: via the Euler-Lotka equation or using the approach in White and Pagano (2007, ref. [54] in the paper)?
}

We estimate $\Rini$ via Euler-Lotka. We have tried to make this clear throughout the main text.

@@

The intrinsic distribution is not our primary focus. 
As we show in Figure 4, the intrinsic distribution overesitmates $\Rini$ as it does not account for spatial effects.
We tried to make clear that our goal is the initial forward distribution.

\rev{
The authors have well demonstrated different sources of sampling bias
arising from contact tracing data, more noticeably, (i) length-bias due
to right-truncation because of retrospective measurement together with
different infection times during the course of the outbreak; (ii) due to the
network structure (e.g. the degree distribution and the degree excess (Line
408). Both sources, the right-truncation and the egocentric point of view
(Line 174), tend to give shorter on average observed intervals. These are
correct and important.
}

\rev{
Any notion of bias is in the context against some "truth". The authors
proposed, and demonstrated by simulation, ways to adjust for the bias in
order to estimate this "truth". However, the authors equate the "true"
distribution to the intrinsic distribution $g(\tau)$ as if "obvious". It is not
obvious and this needs to be proven.
}

We had never equated the true distribution to the intrinsic distribution.
We make clear that the intrinsic distributions are not our goal.

@@

\rev{
I rate this paper very high in scientific importance and in general interest.
In the last 20 years, especially since SARS outbreak in 2003, there has
been a huge volumn of related work, more than those referenced in this
paper. These "intervals" have been widely cited, measured and applied to
outbreak investigations, especially during the early stage. However, there
is a lot of ambiguity. There are occasions that the same terminology is
associated with two different definitions, whereas there are other occasions
that the same definition is assigned to different terminologies by different
researchers.
}

Thank you!

\rev{
I suggest major revision of the paper and hope my comments are
constructive for the revision(s) for better clarification. It will be a
great contribution to this field.
}

We have made major revisions to the manuscript as per suggested comments.
We hope that our manuscript is clearer.

### 2020-05-20 doc/network.tex
changes

\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}\xspace} 
\newcommand{\Ro}{\Rx{0}}
\newcommand{\RR}{\ensuremath{{\mathcal R}}}
\newcommand{\Rini}{\Rx{\textrm{\tiny initial}}}
\newcommand{\Rhat}{\ensuremath{{\hat\RR}}}
\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}

@@

The population-level egocentric kernel is found by integrating the individual-level egocentric over individual variations:
\begin{equation}\label{eq:ego}
\hat{K}(\tau) = \int \hat{k}(\tau; a) f(a) da,
\end{equation}
where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
Essentially, the population-level egocentric kernel accounts for the probability that a susceptible individual has not been infected by the focal individual.
Trapman \textit{et al.} \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
The population-level egocentric generation-interval distribution is:
\begin{equation}
\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s) ds}.
\label{eq:conditional}
\end{equation}
The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized from an \emph{average} infected-susceptible pair; for convenience, we will often omit ``population-level''.
Finally, the link between the growth rate and the egocentric reproductive number are linked by the egocentric generation-interval distribution (and the Euler-Lotka equation) \citep{trapman2016inferring}:
\begin{equation}
\frac{1}{\hat{\RR}} = \int \hat{g}(\tau) \exp(-r \tau) d\tau.
\label{eq:egorR}
\end{equation}
As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution;
this generation-interval-based argument provides a clear biological interpretation for the result presented by \cite{trapman2016inferring}.
% Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.

For example, consider a susceptible-exposed-infected-recovered (SEIR) model, which assumes that latent and infectious periods are exponentially distributed.
The intrinsic generation-interval distribution that corresponds to this model can be written as \citep{Champredon2018equivalence}:
\begin{equation}
g(\tau) = \frac{\sigma \gamma}{\sigma - \gamma} \left(e^{-\gamma \tau} - e^{-\sigma \tau}\right),
\end{equation}
where $1/\sigma$ and $1/\gamma$ are mean latent and infectious periods, respectively.
Assuming a constant per-pair contact rate of $\lambda$ for any pair, we obtain the following egocentric generation-interval distribution:
\begin{equation}
\hat{g}(\tau) = \frac{\sigma (\gamma + \lambda)}{\sigma - (\gamma + \lambda)} \left(e^{-(\gamma + \lambda)\tau} - e^{-\sigma \tau}\right)
\end{equation}
In this case, with fixed infectiousness during the infection period, the effect of accounting for pairwise contacts is the same as an increase in the recovery rate (by the amount of the per-pair contact rate $\lambda$): infecting a susceptible contact is analogous to no longer being infectious (since the contact cannot be infected again); therefore, the resulting egocentric generation-interval distribution is equivalent to the intrinsic generation-interval distribution with mean latent period of $1/\sigma$ and mean infectious period of $1/(\gamma + \lambda)$.
In practice, directly using this distribution to link $r$ and $\RR$ using the Euler-Lotka equation is unrealistic because it requires that we know the per-pair contact rate. 
Instead, the per-pair contact rate can be inferred from the growth rate $r$, assuming that mean and variance of the degree distribution of a network is known (see \cite{trapman2016inferring} supplementary material section 1.4.2);
we briefly describe this relationship in Section~\ref{egosection}.

@@

The egocentric generation interval (\eref{conditional}) only explains some of the reduction in generation intervals that occurs on most networks, however.
Generation intervals are also shortened by indirect connections: a susceptible individual can be infected through another route before the focal individual makes infectious contacts.
Simulations on a small homogeneous network (i.e., complete network) confirm this additional effect (\fref{local}, right panel). 
We can think of simulations on this network as an approximation of (local) infection process in a small household, consisting of 5 individuals;
we expect realistic local network structures (and their effects on the realized generation intervals) to lay between a star network and a complete network.

@@

In a large homogeneously mixing population, the initial forward generation-interval distribution is equivalent to the intrinsic distribution and provides the correct link between the exponential growth rate $r$ and the initial reproduction number \Rini (see \eref{forward}).
In a nonhomogeneous population, the initial forward generation-interval distributions are subject to spatial effects and, therefore, are different from the intrinsic distribution;
since spatial effects have the same effect on how the epidemic spreads as they do on realized generation intervals, we expect the initial forward generation-interval distributions, which implicitly account for the spatial structure, to correctly link $r$ and $\Rini$ through the Euler-Lotka equation (\eref{euler}).

@@

Susceptible depletion happening at all three levels shortens realized generation intervals but acts on different time scales.
Egocentric and local depletion effects are present from the beginning of an epidemic, even when depletion in the global susceptible population is negligible, and can strongly affect the initial spread of an epidemic.
Therefore, we predicted the realized generation intervals during an exponential growth phase to contain information about the contact structure, allowing us to estimate the initial forward generation-interval distribution by simply accounting for the right-truncation.
Simulation studies confirmed our prediction: using the initial forward generation-interval distribution provides the correct link between $r$ and $\RR$.

We compare two methods for estimating the initial forward generation-interval distribution and assume that the initial forward generation-interval distribution follows a gamma distribution.
The gamma approximation of the generation interval distribution has been widely used due to its simplicity \citep{mcbryde2009early, nishiura2009transmission, roberts2011early, trichereau2012estimation, nishiura2015theoretical};
we previously thought that summarizing the entire distribution with two moments (mean and variance) is sufficient to understand the role of generation-interval distributions in linking $r$ and $\RR$ \citep{park2019practical}.
However, further investigation of our methods suggests that making a wrong distributional assumption can lead to biased estimates of the mean and CV of a generation-interval distribution (Appendix A.3), even though the gamma distribution ``looks'' indistinguishable from the shape of the intrinsic generation-interval distribution (derived from the SEIR model).
These results are particularly alarming because it is impossible to know the true shape of the generation-interval distribution for real diseases.
Nonetheless, biases in the parameter estimates of a generation-interval distribution may cancel out and give unbiased estimate of the reproductive number (Appendix A.3).

@@

To study the effects on right-truncation on the observed generation intervals, we use the deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model.
The SEIR model describes how disease spreads in a homogeneously mixing population; it assumes that infected individuals become infectious after a latent period.
We use a \seminar\ model, which extends the SEIR model to have multiple equivalent stages in the latent and infectious periods. This gives latent and infectious periods with Erlang distributions (gamma distributions with integer shape parameters, including the exponential distribution), which are often more realistic approximations than the exponentially distributed periods in the standard SEIR model \citep{anderson1980spread, bailey1964some}: 
\begin{equation}
\begin{aligned}
\frac{dS}{dt} &= - \beta S \frac{\sum_{k=1}^{n_I} I_k}{N}\\
\frac{dE_1}{dt} &= \beta S \frac{\sum_{k=1}^{n_I} I_k}{N} - n_E \sigma E_1\\
\frac{dE_m}{dt} &= n_E \sigma (E_{m-1} - E_m) && \text{for } m = 2, 3, \dots, n_E\\
\frac{dI_1}{dt} &= n_E \sigma E_m - n_I \gamma I_1\\
\frac{dI_n}{dt} &= n_I \gamma (I_{n-1} - I_n) && \text{for } n = 2, 3, \dots, n_I\\
\end{aligned}
\end{equation}
where $S$ is the number of susceptible individuals, $E_m$ is the number of exposed individuals in the $m$-th compartment, $I_n$ is the number of infectious individuals in the $n$-th compartment, and $R$ is the number of recovered individuals.
Parameters of the model are specified as follows: $N$ is the total population size, $\beta$ is the transmission rate, $1/\sigma$ is the mean latent period, $n_E$ is the number of latent compartments, $1/\gamma$ is the mean infectious period, and $n_I$ is the number of infectious compartments.
We show results based on more realistic distributions in Appendix.

@@

\subsection{Estimating generation-interval distributions from contact-tracing data}

The \textbf{population-level method} estimates the initial forward generation-interval distribution by reversing the inverse exponential weighting in the aggregated generation-interval distribution:
\begin{equation}
f_0(\tau) \propto a_0(\tau) \exp(r\tau).
\end{equation}
In order to do so, we first fit a gamma distribution to the observed generation-interval distributions; 
specifically, we estimate the mean $\bar G$ and shape $\alpha$ of a gamma distribution by maximum likelihood.
Then, the effective generation-interval distribution also follows a gamma distribution with mean $\alpha/(\alpha/\bar G - r)$ and shape $\alpha$.

>>>>

\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}\xspace} 
\newcommand{\Ro}{\Rx{0}}
\newcommand{\RR}{\ensuremath{{\mathcal R}}}
\newcommand{\Rini}{\Rx{\textrm{\tiny initial}}}
\newcommand{\Rego}{\Rx{\textrm{\tiny egocentric}}}
\newcommand{\Rhat}{\ensuremath{{\hat\RR}}}
\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}

@@

The population-level egocentric kernel is found by integrating the individual-level egocentric over individual variations:
\begin{equation}\label{eq:ego}
\hat{K}(\tau) = \int \hat{k}(\tau; a) f(a) da,
\end{equation}
where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
Essentially, the population-level egocentric kernel accounts for the probability that a susceptible individual has not been infected by the focal individual.
Trapman \textit{et al.} \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
The population-level egocentric generation-interval distribution is:
\begin{equation}
\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s) ds}.
\label{eq:conditional}
\end{equation}
The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized from an \emph{average} infected-susceptible pair; for convenience, we will often omit ``population-level''.
Finally, the link between the growth rate and the egocentric reproductive number are linked by the egocentric generation-interval distribution (and the Euler-Lotka equation) \citep{trapman2016inferring}:
\begin{equation}
\frac{1}{\Rego} = \int \hat{g}(\tau) \exp(-r \tau) d\tau.
\label{eq:egorR}
\end{equation}
As the egocentric distribution always has a shorter mean than the intrinsic distribution, $\Rego$ will always be smaller than $\Rini$ estimated from the intrinsic distribution;
this generation-interval-based argument provides a clear biological interpretation for the result presented by \cite{trapman2016inferring}.
% Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.

For example, consider a susceptible-exposed-infected-recovered (SEIR) model, which assumes that latent and infectious periods are exponentially distributed.
The intrinsic generation-interval distribution that corresponds to this model can be written as \citep{Champredon2018equivalence}:
\begin{equation}
g(\tau) = \frac{\sigma \gamma}{\sigma - \gamma} \left(e^{-\gamma \tau} - e^{-\sigma \tau}\right),
\end{equation}
where $1/\sigma$ and $1/\gamma$ are mean latent and infectious periods, respectively.
Assuming a constant per-pair contact rate of $\lambda$ for any pair, we obtain the following egocentric generation-interval distribution:
\begin{equation}
\hat{g}(\tau) = \frac{\sigma (\gamma + \lambda)}{\sigma - (\gamma + \lambda)} \left(e^{-(\gamma + \lambda)\tau} - e^{-\sigma \tau}\right)
\end{equation}
In this case, with fixed infectiousness during the infection period, the effect of accounting for pairwise contacts is the same as an increase in the recovery rate (by the amount of the per-pair contact rate $\lambda$). 
Infecting a susceptible contact is analogous to recovery because the contactee cannot be infected again --- the infector cannot no longer transmit infection even if they are infectious (effectively losing infectiousness).
Therefore, the resulting egocentric generation-interval distribution is equivalent to the intrinsic generation-interval distribution with mean latent period of $1/\sigma$ and mean infectious period of $1/(\gamma + \lambda)$.
In practice, directly using the egocentric distribution to link $r$ and $\Rini$ using the Euler-Lotka equation is unrealistic because it requires that we know the per-pair contact rate. 
Instead, the per-pair contact rate can be inferred from the growth rate $r$, assuming that mean and variance of the degree distribution of a network is known (see \cite{trapman2016inferring} supplementary material section 1.4.2);
we briefly describe this relationship in Section~\ref{egosection}.

@@

The egocentric generation interval (\eref{conditional}) only explains some of the reduction in realized generation intervals that occurs on most networks, however.
Generation intervals are also shortened by indirect connections: a susceptible individual can be infected through another route before the focal individual makes infectious contacts.
Simulations on a small homogeneous network (i.e., complete network) confirm this additional effect (\fref{local}, right panel). 
We can think of simulations on this network as an approximation of (local) infection process in a small household, consisting of 5 individuals;
we expect realistic local network structures (and their effects on the realized generation intervals) to lay between a star network and a complete network.

@@

In a large homogeneously mixing population, the initial forward generation-interval distribution is equivalent to the intrinsic distribution and provides the correct link between the exponential growth rate $r$ and the initial reproduction number \Rini (see \eref{forward}).
In a nonhomogeneous population, the initial forward generation-interval distributions are subject to spatial effects and, therefore, are different from the intrinsic distribution;
since spatial effects have the same effect on how the epidemic spreads as they do on realized generation intervals, we predict the initial forward generation-interval distributions, which implicitly account for the spatial structure, to correctly link $r$ and $\Rini$ through the Euler-Lotka equation (\eref{euler}).
Since spatial effects on realized generation intervals are generally expected to be analytically intractable, even in simple networks (e.g., see \cite{tomba2010some} for discussion regarding the realized generation intervals in a household with one infector and two susceptibles), we rely on simulations hereafter to vadliate our prediction.

@@

Susceptible depletion happening at all three levels shortens realized generation intervals but acts on different time scales.
Egocentric and local depletion effects are present from the beginning of an epidemic, even when depletion in the global susceptible population is negligible, and can strongly affect the initial spread of an epidemic.
Therefore, we predicted the realized generation intervals during an exponential growth phase to contain information about the contact structure, allowing us to estimate the initial forward generation-interval distribution by simply accounting for the right-truncation.
Simulation studies confirmed our prediction: using the initial forward generation-interval distribution provides the correct link between $r$ and $\Rini$.

We compare two methods for estimating the initial forward generation-interval distribution and assume that the initial forward generation-interval distribution follows a gamma distribution.
The gamma approximation of the generation interval distribution has been widely used due to its simplicity \citep{mcbryde2009early, nishiura2009transmission, roberts2011early, trichereau2012estimation, nishiura2015theoretical};
we previously thought that summarizing the entire distribution with two moments (mean and variance) is sufficient to understand the role of generation-interval distributions in linking $r$ and $\Rini$ \citep{park2019practical}.
However, further investigation of our methods suggests that making a wrong distributional assumption can lead to biased estimates of the mean and CV of a generation-interval distribution (Appendix A.3), even though the gamma distribution ``looks'' indistinguishable from the shape of the intrinsic generation-interval distribution (derived from the SEIR model).
These results are particularly alarming because it is impossible to know the true shape of the generation-interval distribution for real diseases.
Nonetheless, biases in the parameter estimates of a generation-interval distribution may cancel out and give unbiased estimate of the reproductive number (Appendix A.3).

@@

To study the effects on right-truncation on the observed generation intervals, we use the deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model.
The SEIR model describes how disease spreads in a homogeneously mixing population; it assumes that infected individuals become infectious after a latent period.
We use a \seminar\ model, which extends the SEIR model to have multiple equivalent stages in the latent and infectious periods. This gives latent and infectious periods with Erlang distributions (gamma distributions with integer shape parameters, including the exponential distribution), which are often more flexible than the exponentially distributed periods in the standard SEIR model \citep{anderson1980spread, bailey1964some}: 
\begin{equation}
\begin{aligned}
\frac{dS}{dt} &= - \beta S \frac{\sum_{k=1}^{n_I} I_k}{N}\\
\frac{dE_1}{dt} &= \beta S \frac{\sum_{k=1}^{n_I} I_k}{N} - n_E \sigma E_1\\
\frac{dE_m}{dt} &= n_E \sigma (E_{m-1} - E_m) && \text{for } m = 2, 3, \dots, n_E\\
\frac{dI_1}{dt} &= n_E \sigma E_m - n_I \gamma I_1\\
\frac{dI_n}{dt} &= n_I \gamma (I_{n-1} - I_n) && \text{for } n = 2, 3, \dots, n_I\\
\end{aligned}
\end{equation}
where $S$ is the number of susceptible individuals, $E_m$ is the number of exposed individuals in the $m$-th compartment, $I_n$ is the number of infectious individuals in the $n$-th compartment, and $R$ is the number of recovered individuals.
Parameters of the model are specified as follows: $N$ is the total population size, $\beta$ is the transmission rate, $1/\sigma$ is the mean latent period, $n_E$ is the number of latent compartments, $1/\gamma$ is the mean infectious period, and $n_I$ is the number of infectious compartments.
In the main text, we present results based on exponentially distributed latent and infectious periods;
we show results based on more Erlang distributed latent periods ($n_E=2$), which better matches the incubation period distribution of Ebola virus disease, in Appendix.

@@

\subsection{Estimating the initial forward generation-interval distribution}

The \textbf{population-level method} estimates the initial forward generation-interval distribution by reversing the inverse exponential weighting in the aggregated generation-interval distribution without explicitly accounting for the infection process (i.e., who infected whom) \citep{tomba2010some, nishiura2010time, britton2019estimation}:
\begin{equation}
f_0(\tau) \propto a_0(\tau) \exp(r\tau).
\end{equation}
In order to do so, we first approximate the aggregated distribution $a_0$ with a gamma distribution by assuming that aggregated realized generation intervals (subject to right-truncation) during the exponential growth phase come from the same gamma distribution;
specifically, we estimate the mean $\bar G$ and shape $\alpha$ of a gamma distribution by maximum likelihood.
Then, the initial forward generation-interval distribution also follows a gamma distribution with mean $\alpha/(\alpha/\bar G - r)$ and shape $\alpha$.
We then use the estimated initial forward generation-interval distribution to infer the initial reproductive number $\Rini$ from the estimated growth rate (using the Euler-Lotka equation).
